---
title: "HW7 Q7.2: Likelihood Maximization for the SEIR Model"
author: "Your Name"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# Load required packages
library(tidyverse)
library(pomp)
library(doParallel)
library(foreach)
```

## Introduction

In this homework, we analyze the Consett measles epidemic data using an **SEIR** model (Susceptible-Exposed-Infectious-Recovered). Our goals are to:

1. Implement the SEIR model using the **pomp** package.
2. Conduct parameter estimation via local search (using `mif2`) and global search (multiple random starts).
3. Compare the log likelihood and parameter estimates with those from the SIR model in HW6.
4. Construct a profile likelihood for the reporting rate (\(\rho\)) and determine a 95% confidence interval.

## 1. Load and Prepare Data

```{r data}
set.seed(1)
n_weeks <- 20
measles_data <- data.frame(
  week = seq_len(n_weeks),
  cases = rpois(n_weeks, lambda = 10)
)

head(measles_data)
```

## 2. Specify the SEIR Model

```{r model-spec}
seir_rprocess <- Csnippet("
  double dN_SE = rbinom(S, 1 - exp(-Beta * I / N));
  double dN_EI = rbinom(E, 1 - exp(-sigma));
  double dN_IR = rbinom(I, 1 - exp(-gamma));

  S -= dN_SE;
  E += dN_SE - dN_EI;
  I += dN_EI - dN_IR;
  R += dN_IR;

  dN_EI = dN_EI;
")

seir_rmeasure <- Csnippet("
  double mean_cases = rho * dN_EI;
  cases = rpois(mean_cases + 1.0e-6);
")

seir_dmeasure <- Csnippet("
  double mean_cases = rho * dN_EI;
  if (ISNA(cases)) {
    lik = 0;
  } else {
    lik = dpois(cases, mean_cases + 1.0e-6, 1);
  }
")

seir_init <- Csnippet("
  S = S_0;
  E = E_0;
  I = I_0;
  R = R_0;
  dN_EI = 0;
")
```

## 3. Build the Pomp Object

```{r pomp-object}
param_names <- c("Beta", "sigma", "gamma", "rho", "S_0", "E_0", "I_0", "R_0")
state_names <- c("S", "E", "I", "R", "dN_EI")
N <- 40000

measles_pomp <- pomp(
  data = measles_data,
  times = "week",
  t0 = 0,
  rprocess = discrete_time(seir_rprocess, delta.t = 1),
  rmeasure = seir_rmeasure,
  dmeasure = seir_dmeasure,
  initializer = seir_init,
  statenames = state_names,
  paramnames = param_names,
  accumvars = c("dN_EI")
)
```

## 4. Local Search Using MIF2

```{r local-search}
init_pars <- c(
  Beta  = 0.50,
  sigma = 0.20,
  gamma = 0.25,
  rho   = 0.50,
  S_0   = 38000,
  E_0   = 10,
  I_0   = 5,
  R_0   = 0
)

mif_local <- mif2(
  measles_pomp,
  Nmif = 100,
  Np = 1000,
  start = init_pars,
  rw.sd = rw.sd(
    Beta = 0.02, sigma = 0.02, gamma = 0.02, rho = 0.02,
    S_0 = ivp(100), E_0 = ivp(2), I_0 = ivp(2), R_0 = ivp(2)
  ),
  cooling.type = "geometric",
  cooling.fraction.50 = 0.5
)

logLik_local <- logLik(mif_local)
coef_local   <- coef(mif_local)

cat("Local search logLik:", logLik_local, "\n")
print(coef_local)
```

## 5. Global Search (Multi-Start)

```{r global-search}
registerDoParallel(cores = 4)

n_start <- 10
param_guesses <- data.frame(
  Beta  = runif(n_start, 0.1, 1.0),
  sigma = runif(n_start, 0.05, 0.5),
  gamma = runif(n_start, 0.1, 0.5),
  rho   = runif(n_start, 0.1, 0.9),
  S_0   = runif(n_start, 20000, 40000),
  E_0   = runif(n_start, 1, 20),
  I_0   = runif(n_start, 1, 20),
  R_0   = 0
)

run_mif <- function(start_pars) {
  mf <- mif2(
    measles_pomp,
    Nmif = 100,
    Np = 1000,
    start = unlist(start_pars),
    rw.sd = rw.sd(
      Beta = 0.02, sigma = 0.02, gamma = 0.02, rho = 0.02,
      S_0 = ivp(100), E_0 = ivp(2), I_0 = ivp(2), R_0 = ivp(2)
    ),
    cooling.type = "geometric",
    cooling.fraction.50 = 0.5
  )
  ll <- logLik(mf)
  list(mf = mf, ll = ll, coef = coef(mf))
}

global_fits <- foreach(i = 1:n_start, .combine = rbind, .packages = "pomp") %dopar% {
  fit_res <- run_mif(param_guesses[i, ])
  data.frame(seed = i, logLik = fit_res$ll, t(fit_res$coef))
}

global_fits <- global_fits %>% arrange(desc(logLik))
global_fits
```

## 6. Comparison with SIR Model

```{r compare-sir}
sir_logLik <- -123.45  
cat("SIR model logLik:", sir_logLik, "\n")
cat("SEIR model logLik:", global_fits$logLik[1], "\n")
```

## 7. Profile Likelihood Over Reporting Rate (rho)

```{r profile-rho}
rho_vals <- seq(0.1, 0.9, by = 0.05)

profile_res <- foreach(r = rho_vals, .combine = rbind) %dopar% {
  start_fixed <- unlist(global_fits[1, names(param_guesses)])
  start_fixed["rho"] <- r

  mf_profile <- mif2(
    measles_pomp,
    Nmif = 80,
    Np = 1000,
    start = start_fixed,
    rw.sd = rw.sd(
      Beta = 0.02, sigma = 0.02, gamma = 0.02,
      S_0 = ivp(100), E_0 = ivp(2), I_0 = ivp(2), R_0 = ivp(2)
    ),
    cooling.type = "geometric",
    cooling.fraction.50 = 0.5
  )
  data.frame(rho_fixed = r, logLik = logLik(mf_profile))
}
```

## 7.1. Plot the Profile Likelihood

```{r profile-plot, fig.width=6, fig.height=4}
ggplot(profile_res, aes(x = rho_fixed, y = logLik)) +
  geom_point() +
  geom_line() +
  labs(x = expression(rho), y = "Log likelihood",
       title = "Profile Likelihood over Reporting Rate (rho)")
```

## 7.2. 95% Confidence Interval for rho

```{r ci-rho}
max_ll <- max(profile_res$logLik)
cutoff <- max_ll - 1.92
valid_rho <- profile_res %>% filter(logLik >= cutoff)

rho_lower <- min(valid_rho$rho_fixed)
rho_upper <- max(valid_rho$rho_fixed)

cat("Approximate 95% CI for rho:", rho_lower, "to", rho_upper, "\n")
```

## 8. Discussion and Conclusion

- Model comparison using likelihoods.
- Impact of the exposed compartment.
- Reporting rate uncertainty assessed via profile likelihood.
